% Dates and Times in R
% Computing for Data Analysis
%

```{r, echo=FALSE}
options(width = 50)
```

# Dates and Times in R

R has developed a special representation of dates and times

- Dates are represented by the `Date` class

- Times are represented by the `POSIXct` or the `POSIXlt` class

- Dates are stored internally as the number of days since 1970-01-01

- Times are stored internally as the number of seconds since
  1970-01-01

# Dates in R

Dates are represented by the `Date` class and can be coerced from a
character string using the `as.Date()` function.

```{r}
x <- as.Date("1970-01-01")
x
unclass(x)
unclass(as.Date("1970-01-02"))
```

# Times in R

Times are represented using the `POSIXct` or the `POSIXlt` class

- `POSIXct` is just a very large integer under the hood; it use a
  useful class when you want to store times in something like a data
  frame

- `POSIXlt` is a list underneath and it stores a bunch of other useful
  information like the day of the week, day of the year, month, day of
  the month

There are a number of generic functions that work on dates and times

- `weekdays`: give the day of the week

- `months`: give the month name

- `quarters`: give the quarter number ("Q1", "Q2", "Q3", or "Q4")

# Times in R

Times can be coerced from a character string using the `as.POSIXlt`
or `as.POSIXct` function.

```{r}
x <- Sys.time()
x
p <- as.POSIXlt(x)
names(unclass(p))
p$sec
```

# Times in R

You can also use the `POSIXct` format.

```{r}
x <- Sys.time()
x  ## Already in `POSIXct' format
unclass(x)
x$sec
p <- as.POSIXlt(x)
p$sec
```

# Times in R

Finally, there is the `strptime` function in case your dates are
written in a different format

```{r}
datestring <- c("January 10, 2012 10:40", "December 9, 2011 9:10")
x <- strptime(datestring, "%B %d, %Y %H:%M")
x
class(x)
```

I can *never* remember the formatting strings. Check `?strptime` for
details.

# Operations on Dates and Times

You can use mathematical operations on dates and times. Well, really
just `+` and `-`. You can do comparisons too (i.e. `==`, `<=`)

```{r}
x <- as.Date("2012-01-01")
y <- strptime("9 Jan 2011 11:34:21", "%d %b %Y %H:%M:%S")
x - y
x <- as.POSIXlt(x)
x - y
```

# Operations on Dates and Times

Even keeps track of leap years, leap seconds, daylight savings, and
time zones.

```{r}
x <- as.Date("2012-03-01")
y <- as.Date("2012-02-28")
x - y
x <- as.POSIXct("2012-10-25 01:00:00")
y <- as.POSIXct("2012-10-25 06:00:00", tz = "GMT")
y - x
```

# Summary

- Dates and times have special classes in R that allow for numerical
  and statistical calculations

- Dates use the `Date` class

- Times use the `POSIXct` and `POSIXlt` class

- Character strings can be coerced to Date/Time classes using the
  `strptime` function or the `as.Date`, `as.POSIXlt`, or `as.POSIXct`
